---
title: "Lecture 4: Confidence Intervals - Task Solutions"
author: "Gustave Kenedi"
date: today
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r}
#| echo: false
# Load required packages
library(tidyverse)
library(infer)
library(knitr)

# Set seed for reproducibility
set.seed(42)

# Load ESS France data
ess_france <- read.csv("https://www.dropbox.com/scl/fi/25h4lezq3zuzla94ejmqq/ess_france_11ed.csv?rlkey=lf23ra0i6bvq5dzgtt2u1gy9a&dl=1")

# Clean the data
ess_france_clean <- ess_france |> 
  filter(!stflife %in% c(77, 88))
```

\newpage

# Your Turn #1: Trust in Politicians

**Task:** Compute a 90% CI for the average trust in politicians among French adults using ESS data.

## Setup

```{r}
# Prepare data: filter out refusals (77), don't know (88)
ess_politicians <- ess_france |> 
  filter(!trstplt %in% c(77, 88))

# Check the data
cat("Sample size:", nrow(ess_politicians), "\n")
summary(ess_politicians$trstplt)
```

## Solution

### Step 1: Compute sample mean, SD, and SE

```{r}
summary_stats <- ess_politicians |> 
  summarise(n = n(),
            mean_trust = mean(trstplt),
            sd_trust = sd(trstplt),
            se_trust = sd_trust / sqrt(n))

summary_stats
```

### Step 2: Find the appropriate t critical value

For a 90% CI with $n - 1$ degrees of freedom, we need $t_{0.05, n-1}$:

```{r}
n <- summary_stats$n
t_crit <- qt(0.95, df = n - 1)  # 0.95 = 1 - 0.10/2 for 90% CI
t_crit
```

Note that with $n = `r n`$, the t critical value (`r round(t_crit, 4)`) is close to the z critical value for 90% CI (1.645).

### Step 3: Compute the 90% CI manually

```{r}
x_bar <- summary_stats$mean_trust
se <- summary_stats$se_trust

ci_lower <- x_bar - t_crit * se
ci_upper <- x_bar + t_crit * se

cat("Manual calculation:\n")
cat("90% CI: [", round(ci_lower, 4), ", ", round(ci_upper, 4), "]\n", sep = "")
```

### Step 4: Verify using t.test()

```{r}
t_test_result <- t.test(ess_politicians$trstplt, conf.level = 0.90)
t_test_result$conf.int
```

The results match!

### Step 5: Interpretation

We are **90% confident** that the true average trust in politicians among French adults falls between `r round(ci_lower, 2)` and `r round(ci_upper, 2)` on a scale from 0 (no trust at all) to 10 (complete trust).

\newpage

# Your Turn #2: Interpersonal Trust Across Education Levels

**Task:** Compute 95% confidence intervals for the proportion of people with high interpersonal trust for each education level, and assess whether there are "true" differences across education levels.

## Setup

```{r}
# Filter data and create high trust variable
ess_trust_educ <- ess_france |> 
  filter(!ppltrst %in% c(77, 88),
         education != "") |> 
  mutate(high_trust = ppltrst > 5)

# Check the data
ess_trust_educ |> count(education)
```

## Solution

### Step 1: Compute proportion with high trust for each education level

```{r}
educ_summary <- ess_trust_educ |> 
  summarise(n = n(),
            n_high_trust = sum(high_trust),
            p_hat = mean(high_trust),
            .by = education)

educ_summary
```

### Step 2: Compute standard error for each proportion

```{r}
educ_summary <- educ_summary |> 
  mutate(se = sqrt(p_hat * (1 - p_hat) / n))

educ_summary
```

### Step 3: Compute 95% CIs (Wald interval) for each education level

```{r}
z_crit <- qnorm(0.975)

educ_summary <- educ_summary |> 
  mutate(ci_lower = p_hat - z_crit * se,
         ci_upper = p_hat + z_crit * se)

educ_summary |> select(education, n, p_hat, ci_lower, ci_upper)
```

### Step 4: Visualization

```{r}
#| fig-width: 8
#| fig-height: 4

educ_summary |> 
  ggplot(aes(x = p_hat, y = fct_reorder(education, -p_hat))) +
  geom_errorbar(aes(xmin = ci_lower, xmax = ci_upper),
               width = .1) +
  geom_point(size = 3, color = "#E74C3C") +
  geom_text(aes(label = scales::percent(p_hat, accuracy = 1)),
            vjust = -1, fontface = "bold") +
  scale_x_continuous(labels = scales::percent, limits = c(0, 0.6)) +
  labs(title = "Interpersonal Trust by Education Level in France",
       subtitle = "Point estimates with 95% confidence intervals | ESS Round 11",
       x = "Proportion with high trust (score > 5)",
       y = "Education level") +
  theme_minimal(base_size = 12)
```

### Step 5: Are there "true" differences across education levels?

**Analysis of overlapping CIs:**

```{r}
# Display the CIs side by side for comparison
educ_summary |> 
  select(education, p_hat, ci_lower, ci_upper)
```

**Discussion:**

Looking at the confidence intervals:

1. **Overlap assessment:** We can check whether the CIs overlap. If two CIs do not overlap at all, this suggests a statistically significant difference between those groups.

2. **Pattern observed:** There appears to be a difference in interpersonal trust between those with higher education and those without.

3. **Statistical significance:**
   - If the CIs for the two education groups don't overlap, we have strong evidence of a "true" difference
   - If CIs overlap substantially, we cannot confidently conclude there's a difference

4. **Caution about CI overlap:** Note that even if two CIs overlap slightly, the difference can still be statistically significant. The rule "non-overlapping CIs = significant difference" is conservative. A proper test would be to compute a CI for the *difference* in proportions.

**Conclusion:** Based on the visual pattern and the degree of CI overlap, we can assess whether there is a "true" difference in interpersonal trust between education levels. This could reflect differences in social capital, exposure to diverse perspectives, or other factors associated with educational attainment.

\newpage

# Your Turn #3: Bootstrap vs. CLT-based CIs

**Task:** Compare bootstrap and t-based confidence intervals for mean satisfaction with democracy in the ESS France data.

## Solution

### Step 1: Compute the t-based 95% CI for mean satisfaction with democracy

```{r}
# Summary statistics
mean_stfdem <- mean(ess_france_clean$stfdem)
sd_stfdem <- sd(ess_france_clean$stfdem)
n <- nrow(ess_france_clean)
se_stfdem <- sd_stfdem / sqrt(n)

cat("Sample statistics:\n")
cat("n =", n, "\n")
cat("Mean satisfaction with democracy =", round(mean_stfdem, 2), "\n")
cat("SD =", round(sd_stfdem, 2), "\n")
cat("SE =", round(se_stfdem, 2), "\n")
```

```{r}
# T-based CI
t_ci <- t.test(ess_france_clean$stfdem, conf.level = 0.95)
t_ci$conf.int
```

### Step 2: Compute the bootstrap 95% CI

```{r}
# Bootstrap distribution
set.seed(123)
bootstrap_stfdem <- ess_france_clean |> 
  specify(response = stfdem) |> 
  generate(reps = 1000, type = "bootstrap") |> 
  calculate(stat = "mean")

# Percentile CI
boot_ci_percentile <- bootstrap_stfdem |> 
  get_ci(level = 0.95, type = "percentile")

cat("Bootstrap percentile CI:\n")
boot_ci_percentile
```

```{r}
# SE-based bootstrap CI
boot_ci_se <- bootstrap_stfdem |> 
  get_ci(level = 0.95, type = "se", point_estimate = mean_stfdem)

cat("\nBootstrap SE CI:\n")
boot_ci_se
```

### Step 3: Visualize the bootstrap distribution

```{r}
#| fig-width: 8
#| fig-height: 5

ggplot(bootstrap_stfdem, aes(x = stat)) +
  geom_histogram(bins = 40, fill = "#3498DB", color = "white", alpha = 0.7) +
  geom_vline(xintercept = mean_stfdem, color = "#E74C3C", linewidth = 1.5) +
  geom_vline(xintercept = c(boot_ci_percentile$lower_ci,
                             boot_ci_percentile$upper_ci),
             color = "#9B59B6", linewidth = 1, linetype = "dashed") +
  annotate("text", x = mean_stfdem + 0.05, y = 80,
           label = paste0("Sample mean = ", round(mean_stfdem, 2)),
           hjust = 0, color = "#E74C3C", fontface = "bold") +
  labs(
    title = "Bootstrap Distribution of Mean Satisfaction with Democracy",
    subtitle = "1000 bootstrap samples | Purple dashed lines = 95% CI",
    x = "Bootstrap sample mean (0-10 scale)",
    y = "Count"
  ) +
  theme_minimal(base_size = 12)
```

### Step 4: Compare the two methods

```{r}
comparison <- tibble(
  Method = c("T-based", "Bootstrap (percentile)", "Bootstrap (SE)"),
  Lower = c(t_ci$conf.int[1],
            boot_ci_percentile$lower_ci,
            boot_ci_se$lower_ci),
  Upper = c(t_ci$conf.int[2],
            boot_ci_percentile$upper_ci,
            boot_ci_se$upper_ci),
  Width = Upper - Lower
)

kable(comparison, digits = 3, caption = "Comparison of CI Methods for Mean Satisfaction with Democracy")
```

### Step 5: Discussion

**How similar are they?**

The three confidence intervals are remarkably similar:

- All three methods produce nearly identical lower and upper bounds
- The widths differ by only small amounts on the 0-10 scale

**Why might they differ?**

The methods may differ for several reasons:

1. **Sample size:** With a large sample ($n = `r n`$), the CLT approximation is very accurate, so t-based and bootstrap methods converge

2. **Bootstrap variability:** Each bootstrap run produces slightly different results due to random resampling. With 1000 replicates, this variability is small but present

3. **Method assumptions:**
   - T-based CI assumes the sampling distribution is approximately normal
   - Bootstrap percentile CI makes no distributional assumptions
   - Bootstrap SE CI assumes normality of the bootstrap distribution

**When would they differ more substantially?**

- Small sample sizes ($n < 30$)
- Highly skewed data
- Statistics other than the mean (e.g., median, variance ratio)
- Distributions with heavy tails

\newpage

# Your Turn #4: Trust Across Demographics

**Task:** Investigate whether interpersonal trust (`ppltrst`) differs by education level using ESS France data.

## Solution

### Step 1: Prepare the data

Create education groups and filter the trust variable:

```{r}
# Prepare data: create education groups and filter trust variable
ess_education <- ess_france_clean |> 
  filter(!ppltrst %in% c(77, 88, 99)) |> 
  mutate(
    # eisced: education level (1-7 scale, 55=other)
    education = case_when(
      eisced <= 3 ~ "No higher education",
      eisced >= 4 & eisced <= 7 ~ "Higher education",
      TRUE ~ NA_character_)) |> 
  filter(!is.na(education))

# Check the data
head(ess_education |> select(ppltrst, eisced, education))
ess_education |> count(education)
```

### Step 2: Compute mean trust score for each education group

```{r}
education_summary <- ess_education |> 
  summarise(n = n(),
            mean_trust = mean(ppltrst),
            sd_trust = sd(ppltrst),
            se_trust = sd_trust / sqrt(n),
            .by = education)

education_summary
```

### Step 3: Compute 95% CIs for each group (t-based)

```{r}
# T-based CIs
ci_no_higher <- t.test(ess_education$ppltrst[ess_education$education == "No higher education"])
ci_higher <- t.test(ess_education$ppltrst[ess_education$education == "Higher education"])

cat("No Higher Education:\n")
cat("Mean:", round(ci_no_higher$estimate, 3), "\n")
cat("95% CI: [", round(ci_no_higher$conf.int[1], 3), ", ",
    round(ci_no_higher$conf.int[2], 3), "]\n\n", sep = "")

cat("Higher Education:\n")
cat("Mean:", round(ci_higher$estimate, 3), "\n")
cat("95% CI: [", round(ci_higher$conf.int[1], 3), ", ",
    round(ci_higher$conf.int[2], 3), "]\n", sep = "")
```

### Step 4: Test whether the difference in means is significant using t.test()

```{r}
# Two-sample t-test
two_sample_test <- t.test(ppltrst ~ education, data = ess_education)
two_sample_test
```

**Interpretation of the t-test:**

- The **difference in means** is `r round(diff(two_sample_test$estimate), 3)` (Higher education minus No higher education)
- The **95% CI for the difference** is [`r round(two_sample_test$conf.int[1], 3)`, `r round(two_sample_test$conf.int[2], 3)`]
- `r ifelse(two_sample_test$p.value < 0.05, "Since the CI does not include 0 and p < 0.05, we conclude there is a **statistically significant difference**", "Since the CI includes 0 and p > 0.05, we cannot conclude there is a statistically significant difference")` in interpersonal trust between education groups

### Step 5: Create a visualization comparing the groups with confidence intervals

```{r}
#| fig-width: 8
#| fig-height: 5

# Create summary data for plotting
plot_data <- ess_education |> 
  summarise(mean = mean(ppltrst),
            se = sd(ppltrst) / sqrt(n()),
            lower = mean - 1.96 * se,
            upper = mean + 1.96 * se,
            .by = education)

plot_data |> 
  ggplot(aes(x = education, y = mean, color = education)) +
  geom_point(size = 1) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.05) +
  geom_text(aes(label = round(mean, 2)), vjust = -2, fontface = "bold") +
  scale_color_manual(values = c("No higher education" = "#E74C3C", "Higher education" = "#3498DB")) +
  scale_y_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(
    title = "Interpersonal Trust by Education Level",
    subtitle = "Point estimates with 95% confidence intervals | ESS France Round 11",
    x = "Education Level",
    y = "Mean Trust Score (0-10)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

### Step 6: Use bootstrap to confirm your t-based results

```{r}
# Bootstrap for difference in means
set.seed(999)
boot_diff_edu <- ess_education |> 
  specify(ppltrst ~ education) |> 
  generate(reps = 1000, type = "bootstrap") |> 
  calculate(stat = "diff in means", order = c("Higher education", "No higher education"))

# Bootstrap CI
boot_ci_edu <- boot_diff_edu |> 
  get_ci(level = 0.95, type = "percentile")

cat("Bootstrap 95% CI for difference (Higher - No higher education):\n")
boot_ci_edu
```

```{r}
#| fig-width: 8
#| fig-height: 4

# Visualize bootstrap distribution
ggplot(boot_diff_edu, aes(x = stat)) +
  geom_histogram(bins = 40, fill = "#9B59B6", color = "white", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = c(boot_ci_edu$lower_ci, boot_ci_edu$upper_ci),
             color = "#E74C3C", linewidth = 1.2, linetype = "solid") +
  annotate("text", x = 0.02, y = 85, label = "No difference", hjust = 0) +
  labs(
    title = "Bootstrap Distribution of Difference in Mean Trust",
    subtitle = "Higher education minus No higher education | Red lines = 95% CI | Dashed = 0",
    x = "Difference in means",
    y = "Count"
  ) +
  theme_minimal(base_size = 12)
```

**Comparison of methods:**

```{r}
method_comparison <- tibble(
  Method = c("T-test", "Bootstrap"),
  Lower = c(two_sample_test$conf.int[1], boot_ci_edu$lower_ci),
  Upper = c(two_sample_test$conf.int[2], boot_ci_edu$upper_ci)
)

kable(method_comparison, digits = 3,
      caption = "CI for Difference in Means: T-test vs Bootstrap")
```

Both methods give very similar results. The bootstrap CI confirms the t-test findings about the relationship between education and interpersonal trust.

\newpage

# Your Turn #5: Putting It All Together

**Task:** Design and conduct a complete analysis using ESS France data.

## Example Analysis: Life Satisfaction by Age Group in France

### Research Question

Do younger and older adults in France differ in their average life satisfaction?

### Data Preparation

```{r}
# Create age groups
ess_age_analysis <- ess_france_clean |> 
  mutate(age_group = case_when(
    agea < 35 ~ "Young (18-34)",
    agea >= 35 & agea < 50 ~ "Middle (35-49)",
    agea >= 50 ~ "Older (50+)"),
    age_group = factor(age_group,
                       levels = c("Young (18-34)", "Middle (35-49)", "Older (50+)")))

# Summary by age group
age_summary <- ess_age_analysis |> 
  summarise(n = n(),
            mean_satisfaction = mean(stflife),
            sd_satisfaction = sd(stflife),
            se_satisfaction = sd_satisfaction / sqrt(n),
            lower_ci = mean_satisfaction - 1.96 * se_satisfaction,
            upper_ci = mean_satisfaction + 1.96 * se_satisfaction,
            .by = age_group)

age_summary
```

### CLT-Based Confidence Intervals

```{r}
# Pairwise comparisons using t-tests
young_vs_older <- t.test(
  stflife ~ age_group,
  data = filter(ess_age_analysis, age_group %in% c("Young (18-34)", "Older (50+)"))
)

cat("Young vs. Older adults:\n")
cat("Difference in means:", round(diff(young_vs_older$estimate), 2), "points\n")
cat("95% CI for difference: [",
    round(young_vs_older$conf.int[1], 2), ", ",
    round(young_vs_older$conf.int[2], 2), "]\n", sep = "")
cat("p-value:", format.pval(young_vs_older$p.value, digits = 3), "\n")
```

### Bootstrap Confidence Intervals

```{r}
# Bootstrap for Young vs. Older
set.seed(777)
boot_young_older <- ess_age_analysis |> 
  filter(age_group %in% c("Young (18-34)", "Older (50+)")) |> 
  specify(stflife ~ age_group) |> 
  generate(reps = 1000, type = "bootstrap") |> 
  calculate(stat = "diff in means", order = c("Older (50+)", "Young (18-34)"))

boot_ci_age <- boot_young_older |> 
  get_ci(level = 0.95, type = "percentile")

cat("Bootstrap 95% CI for difference (Older - Young):\n")
boot_ci_age
```

### Visualization

```{r}
#| fig-width: 9
#| fig-height: 5

# Main plot: CIs by age group
ggplot(age_summary, aes(x = age_group, y = mean_satisfaction, color = age_group)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci),
                width = 0.05) +
  scale_color_manual(values = c("#3498DB", "#9B59B6", "#E74C3C")) +
  scale_y_continuous(limits = c(0, 10), breaks = 0:10) +
  labs(
    title = "Life Satisfaction by Age Group in France",
    subtitle = "Point estimates with 95% confidence intervals | ESS Round 11",
    x = "",
    y = "Average life satisfaction (0-10)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

```{r}
#| fig-width: 8
#| fig-height: 4

# Bootstrap distribution for Young vs. Older difference
ggplot(boot_young_older, aes(x = stat)) +
  geom_histogram(bins = 40, fill = "#2D3E50", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = c(boot_ci_age$lower_ci, boot_ci_age$upper_ci),
             color = "#E74C3C", linewidth = 1.2) +
  labs(
    title = "Bootstrap Distribution: Difference in Life Satisfaction",
    subtitle = "Older (50+) minus Young (18-34) | Red lines = 95% CI",
    x = "Difference in mean satisfaction",
    y = "Count"
  ) +
  theme_minimal(base_size = 11)
```

### Interpretation

Based on our analysis of European Social Survey data for France (Round 11, 2023-2024):

**Key Findings:**

1. **Point estimates:** Young adults (18-34) average approximately `r round(age_summary$mean_satisfaction[1], 1)` on life satisfaction, while older adults (50+) average `r round(age_summary$mean_satisfaction[3], 1)` on the 0-10 scale.

2. **Confidence intervals:** All age groups show moderate to high life satisfaction. The 95% CIs for all groups are:
   - Young (18-34): [`r round(age_summary$lower_ci[1], 1)`, `r round(age_summary$upper_ci[1], 1)`]
   - Middle (35-49): [`r round(age_summary$lower_ci[2], 1)`, `r round(age_summary$upper_ci[2], 1)`]
   - Older (50+): [`r round(age_summary$lower_ci[3], 1)`, `r round(age_summary$upper_ci[3], 1)`]

3. **Statistical significance:** The difference between young and older adults `r ifelse(young_vs_older$p.value < 0.05, "is", "is not")` statistically significant at the 5% level (p = `r format.pval(young_vs_older$p.value, digits = 3)`).

4. **Bootstrap confirmation:** The bootstrap 95% CI for the difference [`r round(boot_ci_age$lower_ci, 2)`, `r round(boot_ci_age$upper_ci, 2)`] `r ifelse(boot_ci_age$lower_ci > 0 | boot_ci_age$upper_ci < 0, "does not include", "includes")` zero, `r ifelse(boot_ci_age$lower_ci > 0 | boot_ci_age$upper_ci < 0, "confirming", "consistent with")` the t-test result.

**Limitations:**

- This analysis uses cross-sectional data; we cannot distinguish between age effects and cohort effects
- Life satisfaction is self-reported and may be subject to social desirability bias
- The ESS sample may not be perfectly representative of all French adults
- Other confounding variables (income, health, employment) are not controlled for
