---
title: "Lecture 5: Hypothesis Testing - Exercise Solutions"
subtitle: "L2 - Statistics"
author: "Gustave Kenedi"
date: today
format:
  pdf:
    toc: true
    toc-depth: 1
    number-sections: true
    colorlinks: true
    geometry: margin=2.5cm
editor: source
---

```{r}
#| echo: false
#| output: false
#| message: false
#| warning: false

# Load required packages
library(tidyverse)
library(moderndive)
library(infer)
library(knitr)
library(scales)
library(effectsize)

# Set seed for reproducibility
set.seed(42)

# Load ESS data
ess_france <- read.csv("https://www.dropbox.com/scl/fi/25h4lezq3zuzla94ejmqq/ess_france_11ed.csv?rlkey=lf23ra0i6bvq5dzgtt2u1gy9a&dl=1")
```

\newpage

# Your Turn 1: Hypothesis Test Overview

**Scenario:** A government report states that French adults have an average interest in politics of **5** on a 0 (*Not at all interested*) – 10 (*Very interested*) scale. You survey a random sample of 50 French university students and find: $\bar{x} = 5.8$, with standard deviation $s = 2.3$.

## Step 1: State the hypotheses

$$H_0: \mu = 5 \quad \text{(students have the same average political interest as the general population)}$$
$$H_A: \mu \neq 5 \quad \text{(students' average political interest differs from the general population)}$$

We use a **two-sided** alternative because we have no prior reason to expect students to be either more or less politically interested, they could go either way.

## Step 2: Choose $\alpha$

$$\alpha = 0.05$$

We accept a 5% probability of incorrectly rejecting $H_0$ (Type I error).

## Step 3: Compute the test statistic

$$t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} = \frac{5.8 - 5}{2.3 / \sqrt{50}} = \frac{0.8}{0.3253} \approx 2.46$$

```{r}
# Given values
n     <- 50
x_bar <- 5.8
s     <- 2.3
mu0   <- 5

# Standard error
se <- s / sqrt(n)
se

# Test statistic
t_stat <- (x_bar - mu0) / se
t_stat
```

## Step 4: Compute the p-value

The p-value is $P(|\bar{X} - 5| \geq 0.8 \mid H_0)$, i.e., the probability of observing a sample mean at least 0.8 units away from 5 if the null hypothesis were true.

```{r}
# Two-sided p-value using t distribution with df = n - 1
p_value <- 2 * pt(-abs(t_stat), df = n - 1)
p_value
```

$$p\text{-value} = 2 \times P(T_{49} \leq -2.46) \approx 0.017$$

## Step 5: Make a decision

Since $p\text{-value} \approx 0.017 < \alpha = 0.05$, we **reject $H_0$ at the 5% level**.

## Step 6: Interpret in context

There is statistically significant evidence ($t = 2.46$, $p = 0.017$) that French university students have a higher average political interest score ($\bar{x} = 5.8$) than the general adult population ($\mu_0 = 5$). The difference of 0.8 points is unlikely to be due to sampling variation alone.

\newpage

# Your Turn 2: One-Sample Tests

```{r}
library(tidyverse)

ess_france <- read.csv("https://www.dropbox.com/scl/fi/25h4lezq3zuzla94ejmqq/ess_france_11ed.csv?rlkey=lf23ra0i6bvq5dzgtt2u1gy9a&dl=1")

ess_stflife <- ess_france |>
  filter(stflife %in% 0:10)
```

## Question 1: State the hypotheses

$$H_0: \mu = 6.5 \quad \text{(average life satisfaction equals 6.5)}$$
$$H_A: \mu \neq 6.5 \quad \text{(average life satisfaction differs from 6.5)}$$

We use a two-sided alternative as there is no prior reason to expect satisfaction to be specifically above or below 6.5.

## Question 2: Manually compute the test statistic and p-value

```{r}
# Descriptive statistics
n    <- nrow(ess_stflife)
n

xbar <- mean(ess_stflife$stflife)
xbar

s    <- sd(ess_stflife$stflife)
s

mu0 <- 6.5

# Test statistic
se     <- s / sqrt(n)
t_stat <- (xbar - mu0) / se
t_stat

# Two-sided p-value
p_val <- 2 * pt(-abs(t_stat), df = n - 1)
p_val
```

$$t = \frac{\bar{x} - 6.5}{s/\sqrt{n}} = \frac{`r round(xbar, 2)` - 6.5}{`r round(s, 2)`/\sqrt{`r n`}} \approx `r round(t_stat, 2)`$$

## Question 3: Verify with `t.test()`

```{r}
t.test(ess_stflife$stflife, mu = 6.5)
```

The `t.test()` output matches our manual computation exactly.

## Question 4: Decision and interpretation

Since $p$-value $= `r format.pval(p_val, digits = 3)` < \alpha = 0.05$, we **reject $H_0$** at the 5% level.

There is strong evidence that average life satisfaction among French adults differs from 6.5. The observed mean ($\bar{x} = `r round(xbar, 2)`$) is statistically significantly higher than 6.5.

## Question 5: 95% confidence interval

The 95% CI is $\bar{x} \pm t^*_{n-1} \cdot \dfrac{s}{\sqrt{n}}$, where $t^*_{n-1}$ is the 97.5th percentile of the $t_{n-1}$ distribution.

```{r}
# Manual construction
crit_value  <- qt(0.975, df = n - 1)
crit_value

ci_low  <- xbar - crit_value * se
ci_high <- xbar + crit_value * se

c(ci_low, ci_high)
```

$$\left[\bar{x} - t^*_{n-1} \cdot \frac{s}{\sqrt{n}},\ \bar{x} + t^*_{n-1} \cdot \frac{s}{\sqrt{n}}\right] = \left[`r round(ci_low, 2)`,\ `r round(ci_high, 2)`\right]$$

**Verify with `t.test()`:**

```{r}
ci <- t.test(ess_stflife$stflife, mu = 6.5)$conf.int
ci
```

The 95% CI is $[`r round(ci[1], 2)`,\ `r round(ci[2], 2)`]$, which **does not contain** $\mu_0 = 6.5$. This is fully consistent with our decision to reject $H_0$. As established in the lecture, a two-sided test at level $\alpha$ rejects $H_0$ if and only if $\mu_0$ falls outside the $(1-\alpha)$ CI.

\newpage

# Your Turn 3: Two-Sample Tests

## Part 1: Life satisfaction by gender

### Question 1: Summary statistics

```{r}
ess_gender <- ess_france |>
  filter(stflife %in% 0:10, gndr %in% 1:2) |>
  mutate(gender = recode(gndr, `1` = "Male", `2` = "Female"))

gender_summary <- ess_gender |>
  group_by(gender) |>
  summarise(
    n    = n(),
    mean = mean(stflife),
    sd   = sd(stflife),
    se   = sd / sqrt(n)
  )

kable(gender_summary, digits = 3)
```

### Question 2: 95% CIs for both means

```{r}
gender_ci <- gender_summary |>
  mutate(
    ci_lower = mean - qt(0.975, df = n - 1) * se,
    ci_upper = mean + qt(0.975, df = n - 1) * se
  )

kable(gender_ci |> select(gender, mean, ci_lower, ci_upper), digits = 3)
```

The 95% CIs for males ($[`r round(gender_ci$ci_lower[gender_ci$gender=="Male"],2)`, `r round(gender_ci$ci_upper[gender_ci$gender=="Male"],2)`]$) and females ($[`r round(gender_ci$ci_lower[gender_ci$gender=="Female"],2)`, `r round(gender_ci$ci_upper[gender_ci$gender=="Female"],2)`]$) **overlap substantially** — already suggesting no significant difference.

### Question 3: Two-sample t-test

**Manual computation (Welch's t-test):**

```{r}
n1    <- gender_summary$n[gender_summary$gender == "Female"]
xbar1 <- gender_summary$mean[gender_summary$gender == "Female"]
s1    <- gender_summary$sd[gender_summary$gender == "Female"]

n2    <- gender_summary$n[gender_summary$gender == "Male"]
xbar2 <- gender_summary$mean[gender_summary$gender == "Male"]
s2    <- gender_summary$sd[gender_summary$gender == "Male"]

SE_gender <- sqrt(s1^2/n1 + s2^2/n2)
SE_gender

t_manual  <- (xbar1 - xbar2) / SE_gender
t_manual

p_value <- 2 * pt(-abs(t_manual), df = 1730.8)
p_value
```

**Verify with `t.test()`:**

```{r}
t_gender <- t.test(stflife ~ gender, data = ess_gender)
t_gender
```

### Question 4: Interpretation

- **Statistically significant?** No — $p = `r round(t_gender$p.value, 3)` > 0.05$. We fail to reject $H_0$.
- **At what level?** The result is not significant even at $\alpha = 0.10$.
- **Practically meaningful?** The mean difference is only `r abs(round(diff(t_gender$estimate), 2))` points on a 0–10 scale — negligible.

**Conclusion:** There is no evidence of a difference in average life satisfaction between men and women in France.

## Part 2: Very satisfied with democracy by education

### Question 1: 95% CIs for both proportions

```{r}
ess_dem_educ <- ess_france |>
  filter(stfdem %in% 0:10) |>
  mutate(
    very_sat    = stfdem >= 8,
    higher_educ = case_when(education == "High school diploma or less" ~ "No higher education",
                                 education != "High school diploma or less" ~ "Higher education")
  )

prop_summary <- ess_dem_educ |>
  group_by(higher_educ) |>
  summarise(
    n       = n(),
    x       = sum(very_sat),
    p_hat   = mean(very_sat),
    ci_lower = prop.test(x, n)$conf.int[1],
    ci_upper = prop.test(x, n)$conf.int[2]
  )

kable(prop_summary |> select(higher_educ, n, p_hat, ci_lower, ci_upper), digits = 3)
```

The CIs **do not overlap** — strong visual evidence of a difference before even running the test.

### Question 2: Two-sample proportion test

**Manual computation:**

```{r}
x1 <- prop_summary$x[prop_summary$higher_educ == "Higher education"]
n1 <- prop_summary$n[prop_summary$higher_educ == "Higher education"]
x2 <- prop_summary$x[prop_summary$higher_educ == "No higher education"]
n2 <- prop_summary$n[prop_summary$higher_educ == "No higher education"]

p_hat1   <- x1 / n1
p_hat2   <- x2 / n2
p_pooled <- (x1 + x2) / (n1 + n2)

SE_prop <- sqrt(p_pooled * (1 - p_pooled) * (1/n1 + 1/n2))
SE_prop

z_prop  <- (p_hat1 - p_hat2) / SE_prop
z_prop

p_prop  <- 2 * pnorm(-abs(z_prop))
p_prop
```

**Verify with `prop.test()`:**

```{r}
prop.test(x = c(x1, x2), n = c(n1, n2), correct = FALSE)
```

### Question 3: Interpretation

- **Statistically significant?** No — $p `r ifelse(p_prop < 0.001, "< 0.001", paste0("= ", round(p_prop,3)))` \gg 0.05$. We fail to reject $H_0$.
- **Practically meaningful?** The proportions are `r round(p_hat1*100,1)`% (higher education) vs `r round(p_hat2*100,1)`% (no higher education) — a difference of `r round((p_hat1-p_hat2)*100,1)` percentage points, which turns out to not be substantively meaningful.

**Conclusion:** Those with higher education are not significantly more likely to be very satisfied with democracy.

\newpage

# Your Turn 4: Paired Tests

## Question 1: Create the difference variable

```{r}
ess_eco_gov <- ess_france |>
  filter(stfeco %in% 0:10, stfgov %in% 0:10) |>
  mutate(diff = stfeco - stfgov)

nrow(ess_eco_gov)
mean(ess_eco_gov$stfeco)
mean(ess_eco_gov$stfgov)
mean(ess_eco_gov$diff)
```

## Question 2: Paired t-test

**Manual computation** (paired t-test = one-sample t-test on the differences):

```{r}
d_bar  <- mean(ess_eco_gov$diff)
d_bar

s_d    <- sd(ess_eco_gov$diff)
s_d

n_pair <- nrow(ess_eco_gov)
n_pair

se_d    <- s_d / sqrt(n_pair)
t_pair  <- d_bar / se_d
t_pair
p_pair  <- 2 * pt(-abs(t_pair), df = n_pair - 1)
p_pair
```

$$t = \frac{\bar{d}}{s_d/\sqrt{n}} = \frac{`r round(d_bar,3)`}{`r round(s_d,3)`/\sqrt{`r n_pair`}} \approx `r round(t_pair, 2)`$$

**Verify with `t.test()`:**

```{r}
t.test(ess_eco_gov$stfeco, ess_eco_gov$stfgov, paired = TRUE)
```

## Question 3: Interpretation

Since $p `r ifelse(p_pair < 0.001, "< 0.001", paste0("\\approx ", round(p_pair,3)))` > 0.05$ we **fail to reject $H_0$**. French adults are not significantly more satisfied with the **economy** than with the **government** (mean difference $\bar{d} = `r round(d_bar, 2)`$, 95% CI: $[`r round(t.test(ess_eco_gov$stfeco, ess_eco_gov$stfgov, paired=TRUE)$conf.int[1], 2)`,\ `r round(t.test(ess_eco_gov$stfeco, ess_eco_gov$stfgov, paired=TRUE)$conf.int[2], 2)`]$, $p > 0.05$).

\newpage

# Your Turn 5: Chi-Square Tests

## Q1: Contingency table

```{r}
ess_lifesat_cat <- ess_france |>
  filter(stflife %in% 0:10, gndr %in% 1:2) |>
  mutate(gender    = recode(gndr, `1` = "Male", `2` = "Female"),
         sat_level = case_when(stflife <= 4  ~ "Low (0-4)",
                               stflife <= 7  ~ "Medium (5-7)",
                               stflife <= 10 ~ "High (8-10)"))

cont_table_gs <- table(ess_lifesat_cat$gender, ess_lifesat_cat$sat_level)
addmargins(cont_table_gs)
```

## Q2: Expected frequencies

```{r}
E_gs <- chisq.test(cont_table_gs)$expected
E_gs
```

## Q3: Chi-square statistic (manual)

```{r}
O_gs      <- as.matrix(cont_table_gs)
chi_man   <- sum((O_gs - E_gs)^2 / E_gs)
df_gs     <- (nrow(O_gs) - 1) * (ncol(O_gs) - 1)

chi_man
df_gs
```

## Q4: p-value

```{r}
p_gs <- pchisq(chi_man, df = df_gs, lower.tail = FALSE)
p_gs
```

## Q5: Verify with `chisq.test()`

```{r}
chisq.test(cont_table_gs)
```

## Q6: Interpretation

```{r}
#| echo: false
props_gs <- ess_lifesat_cat |>
  group_by(gender, sat_level) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(prop = n / sum(n), .by = gender) |>
  mutate(sat_level = factor(sat_level, levels = c("Low (0-4)", "Medium (5-7)", "High (8-10)")))

ggplot(props_gs, aes(x = gender, y = prop, fill = sat_level)) +
  geom_bar(stat = "identity", position = "fill", color = "white", linewidth = 0.3) +
  geom_text(aes(label = percent(prop, accuracy = 0.1)),
            position = position_fill(vjust = 0.5),
            color = "white", fontface = "bold", size = 5) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("Low (0-4)"    = "#E74C3C",
                               "Medium (5-7)" = "#F39C12",
                               "High (8-10)"  = "#27AE60")) +
  labs(title = "Life satisfaction categories by gender",
       subtitle = paste0("χ² = ", round(chi_man, 2),
                        ", df = ", df_gs,
                        ", p = ", format.pval(p_gs, digits = 3, eps = 0.001)),
       x = "Gender", y = "Proportion", fill = "Life satisfaction") +
  theme(legend.position = "bottom")
```

We **fail to reject $H_0$** (p = `r format.pval(p_gs, digits = 3, eps = 0.001)` > 0.05). There is no statistically significant evidence of an association between gender and life satisfaction categories.

\newpage

# Your Turn 6: Randomization Tests

## Q1: Observed difference

```{r}
ess_gender_infer <- ess_france |>
  filter(stflife %in% 0:10, gndr %in% 1:2) |>
  mutate(gender = recode(gndr, `1` = "Male", `2` = "Female"))

obs_diff <- ess_gender_infer |>
  specify(stflife ~ gender) |>
  calculate(stat = "diff in means", order = c("Male", "Female"))

cat("Observed difference (Male - Female):", round(obs_diff$stat, 3), "\n")
```

## Q2: Generate 10,000 permutations

```{r}
set.seed(2024)
null_dist <- ess_gender_infer |>
  specify(stflife ~ gender) |>
  hypothesize(null = "independence") |>
  generate(reps = 10000, type = "permute") |>
  calculate(stat = "diff in means", order = c("Male", "Female"))
```

## Q3: Visualize and calculate p-value

```{r}
null_dist |>
  visualize() +
  shade_p_value(obs_stat = obs_diff, direction = "both") +
  labs(title = "Permutation distribution: gender difference in life satisfaction",
       x = "Difference in mean life satisfaction (Male - Female)",
       y = "Count")
```

```{r}
p_perm <- null_dist |>
  get_p_value(obs_stat = obs_diff, direction = "both")

cat("Permutation p-value:", p_perm$p_value, "\n")
```

## Q4: Compare to theory-based t-test

```{r}
p_theory <- t.test(stflife ~ gender, data = ess_gender_infer)$p.value
cat("Theory-based p-value:", round(p_theory, 4), "\n")
```

Both approaches give similar p-values (`r round(p_perm$p_value, 3)` vs `r round(p_theory, 3)`) and reach the same conclusion: **no evidence of a gender difference in life satisfaction**. The permutation test confirms that the t-test result is not an artefact of its normality assumptions.

---

# Your Turn 7: Power and Sample Size

## Part 1: Compute power

**Setup:** $\sigma = 2$, $\mu_0 = 7.0$, $\mu_1 = 6.8$, $n = 80$, $\alpha = 0.05$

### Q1: Manual computation

```{r}
mu0   <- 7.0
mu1   <- 6.8
sigma <- 2.0
n     <- 80
alpha <- 0.05

se   <- sigma / sqrt(n)
crit <- mu0 + qt(1 - alpha / 2, df = n - 1) * se   # upper critical value

# Power = P(X-bar < lower crit | mu = mu1)  (effect is negative: mu1 < mu0)
crit_lo <- mu0 - qt(1 - alpha / 2, df = n - 1) * se

power_manual <- pnorm(crit_lo, mean = mu1, sd = se, lower.tail = TRUE)

cat("SE =", round(se, 4), "\n")
cat("Lower critical value =", round(crit_lo, 4), "\n")
cat("Power =", round(power_manual, 3), "\n")
```

Since $\mu_1 = 6.8 < \mu_0 = 7.0$, the relevant rejection region is the **lower** tail — we reject when $\bar{X}$ is sufficiently *below* 7.0.

### Q2: Verify with `power.t.test()`

```{r}
power.t.test(delta = abs(mu1 - mu0), sd = sigma, n = n,
             sig.level = alpha, type = "one.sample")
```

With $n = 80$ the power is modest (~60%) — this study is underpowered to reliably detect a 0.2-point difference.

---

## Part 2: Minimum sample size

**Setup:** detect $|\mu_1 - \mu_0| = 0.4$, $\sigma = 2$, power = 0.80, $\alpha = 0.05$

### Q1: Manual computation

```{r}
delta  <- 0.4    # minimum effect to detect
sigma2 <- 2.0
power_target <- 0.80

z_alpha2 <- qnorm(1 - alpha / 2)       # z_{alpha/2} ≈ 1.96
z_beta   <- qnorm(power_target)         # z_{1-beta}  ≈ 0.84

n_manual <- ((z_alpha2 + z_beta) / (delta / sigma2))^2

cat("z_alpha/2 =", round(z_alpha2, 4), "\n")
cat("z_beta    =", round(z_beta,   4), "\n")
cat("n (formula) =", round(n_manual, 1), "→ round up to", ceiling(n_manual), "\n")
```

### Q2: Verify with `power.t.test()`

```{r}
power.t.test(delta = delta, sd = sigma2, power = power_target,
             sig.level = alpha, type = "one.sample")
```

The formula gives $n \approx$ `r ceiling(n_manual)` while `power.t.test()` returns a slightly larger value due to using the exact non-central t-distribution rather than the normal approximation. Always round up and use `power.t.test()` for the final answer.